---
title: 'Reproducible Research: Peer Assessment 1'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---




## Loading and preprocessing the data
This report takes a raw data set of activity data from wearable computing. More data 
can be found in the README accompanying at the following Github repository link
<https://github.com/domlloyd/RepData_PeerAssessment1>

Before running the report we will need to source the various R libraries used to produce the report

```{r}
# This script uses:
# - the "readr" library for fast reading of files using read_table
# - the "dplyr" library for summarising the data
# - the "stringr" library to use str_pad to pad the time strings
library("readr", quietly = TRUE, warn.conflicts = FALSE)
library("dplyr", quietly = TRUE, warn.conflicts = FALSE)
library("stringr", quietly = TRUE, warn.conflicts = FALSE)
library("ggplot2", quietly = TRUE, warn.conflicts = FALSE)
```

And then we read the file of activity data, getting rid of NA values. 

```{r}
# the raw data has been unzipped to our RProject directory
# We setwd to the project directory we have forked. 
# NB If running this program make sure the raw data file
# household_power_consumption.txt hs been unzipped and put in to the
# working directory shown below.

setwd("..//RepData_PeerAssessment1")

# clear the workspace - we do this for ease of seeing which variables have been generated. 
rm(list=ls())

# read the activity file
activity_all <- read.table("activity//activity.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE,
                           na.strings = "NA")

# filter out NA values for steps
activity_all_valid <- filter(activity_all, !is.na(steps))
activity_all_valid2 <- filter(activity_all, !is.na(steps))
activity_all_na    <- filter(activity_all, is.na(steps))
# pad the time taken from interval field
activity_all_valid$time <- str_pad(as.character(activity_all_valid$interval), width = 4, side = "left", pad = "0")
#activity_all_valid$datetime <- strptime(paste(activity_all_valid$date, activity_all_valid$time), '%Y-%m-%d %H%M')

```





## What is mean total number of steps taken per day?

First we will build a data table with the total number of steps per day

``` {r}
# work out the total number of steps per day
total_steps_per_day <- summarise(group_by(activity_all_valid, date),
                                 sum_steps = sum(steps))

```


Then we can plot a histogram of the total number of steps over the activity days.

```{r, echo=FALSE}
# and plot the total number of steps per day on a histogram
x <- total_steps_per_day$sum_steps
xlab <- ""
# make sure to convert the date to POSIXlt
y <- strptime(total_steps_per_day$date, '%Y-%m-%d')
ylab <- "Total number of steps"
main_title <- "Total number of steps per day"
plot(y, x, type = 'h',xlab = xlab, ylab = ylab, main = main_title)
```


The mean number of steps per day is: 

``` {r, echo=F}
round(mean(total_steps_per_day$sum_steps), digits = 0)
```

and the median number of steps is: 
``` {r,echo=F}
median(total_steps_per_day$sum_steps)
```


## What is the average daily activity pattern?
The following graph shows the average daily activity pattern by taking the averages across all days for the five minute periods. 
```{r, echo=FALSE}
# summarise  by the time segment to get the average daily activity pattern
steps <- as.numeric(activity_all_valid$steps)
times <- as.character(activity_all_valid$time)
new_activity_all_valid = cbind.data.frame(steps,times)
summary_adap <- summarise(group_by(new_activity_all_valid, times), mean_steps = mean(steps))
plot(as.character(summary_adap$times), as.numeric(summary_adap$mean_steps), type = "l", xlab = "time of day", ylab = "average number of steps")
```

From the graph we can see the maximum number of steps occurring in a 5 minute interval in an average day is: 
```{r, echo=FALSE}
max(summary_adap$mean_steps)
max_steps_in_adap <- max(summary_adap$mean_steps)
```
And this peak number of steps occurs at
```{r, echo=FALSE}
filter(summary_adap, mean_steps == max_steps_in_adap)
```



## Imputing missing values

We have come up with a method for filling in NA values with the average number of steps at interval

```{r}
averages <- aggregate(x=list(steps=activity_all$steps), by=list(interval=activity_all$interval),
                      FUN=mean, na.rm=TRUE)
merged <- merge.data.frame(activity_all_na,averages, by = c("interval"))
activity_all_na$steps <- merged$steps.y
final_data = rbind.data.frame(activity_all_na,activity_all_valid2)
```

Redo the sums from the original set 0f graphs. 

``` {r}
# work out the total number of steps per day
total_steps_per_day2 <- summarise(group_by(final_data, date),
                                 sum_steps = sum(steps))

```


Then we can plot a histogram of the total number of steps over the activity days.

```{r, echo=FALSE}
# and plot the total number of steps per day on a histogram
x <- total_steps_per_day2$sum_steps
xlab <- ""
# make sure to convert the date to POSIXlt
y <- strptime(total_steps_per_day2$date, '%Y-%m-%d')
ylab <- "Total number of steps"
main_title <- "Total number of steps per day"
plot(y, x, type = 'h',xlab = xlab, ylab = ylab, main = main_title)
```


The mean number of steps per day is: 

``` {r, echo=F}
round(mean(total_steps_per_day2$sum_steps), digits = 0)
```

and the median number of steps is: 
``` {r,echo=F}
median(total_steps_per_day2$sum_steps)
```
The median number of steps has gone higher because we have more values taken in to consideration in the calculation. 



## Are there differences in activity patterns between weekdays and weekends?

```{r}
final_data$day <- weekdays(strptime(final_data$date,'%Y-%m-%d'))
weekday_weekend <- function(day) {
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("not a valid day above")
}
final_data$weekend <- sapply(final_data$day, FUN=weekday_weekend)


```
  
```{r, echo=F}

# plot graph
steps <- as.numeric(final_data$steps)
interval <- as.character(final_data$interval)
daygroup <- as.character(final_data$weekend)
final_activity_all_valid = cbind.data.frame(steps,interval,daygroup)
summary_adap_weekends <- summarise(group_by(final_activity_all_valid, interval, daygroup), meanstep = mean(steps))

ggplot(data=summary_adap_weekends, aes(x=interval, y=meanstep)) + geom_line(aes(group=daygroup)) + facet_grid(daygroup~.) + xlab("5 minute interval") + ylab("average number of steps")



```  
  